\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2016-03-07 write_up/cricket.tex
added an abstract + abstract v. of conditioning factors

@@

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches--- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.

@@

The data suggest that winning the toss has a sizable impact on the probability of winning, especially in closely contested games. The data also suggest that the advantage varies considerably and systematically in expected ways, with advantage being greater in  day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match was played between closely matched teams.

>>>>

\begin{center}\textbf{Abstract}\\\end{center}
In a competitive sport, every little thing matters. Yet many sports leave some large levers out of the reach of the teams, in the hands of fate. In cricket, world's second most popular sport by some measures \footnote{See \citet{Economist2011} in The Economist.}, one such lever---the toss---has been subject to much recent attention. Using a large novel dataset of 44,224 cricket matches, we estimate the impact of winning the toss on the probability of winning. The data suggest that winning the toss increases the chance of winning by a small ($\sim$ 2.8\%) but significant margin. The advantage varies heftily and systematically, by how closely matched the competing teams are, and by playing conditions---tautologically, winning the toss in conditions where the toss grants a greater advantage, for e.g., in day and night matches, has a larger impact on the probability of winning.  
\clearpage

@@

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches--- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically. We next assess whether the advantage of winning the toss varies, broadly speaking, by how closely matched the teams are, and by how large an advantage winning the toss grants---the advantage of winning the toss is greater in certain playing conditions than others. We find that the advantage of winning the toss varies widely and systematically, in expected ways.

@@

The data suggest that winning the toss has a sizable impact on the probability of winning, especially in closely contested games. The data also suggest that the advantage varies considerably and systematically ---in expected ways---, with advantage greater in day and night matches, matches in which Duckworth-Lewis is used to adjust scores, and where the match is played between closely matched teams. In showing so, the data lend credence to, and quantify, the suspicion that many of the cricket fans have long had---that tosses matter. Besides that, the data also help quantify the bias in Duckworth-Lewis method. More generally, the analysis we do here could be replicated elsewhere to assess bias in competing methods, and used to prove that a particular method is better or worse than the Duckworth-Lewis.
